Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-10 text/ms.tex
Edit edit edit

Black swans are statistically improbable events that nonetheless occur --- often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

% TODO: cite panels c, d in Fig 1

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$. We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}. For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts (``random effects'') at the taxonomic class, order, and species level (Supporting Information). We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}b).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}). However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}). For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example.)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans were usually downward events and were detected more frequently in longer time series and in populations with a smaller magnitude of process noise. Black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black-swan events were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors and were discarded (Supporting Information). A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. As an example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}. Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent). Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Information).
}
\label{fig:correlates}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{
Posterior probability distributions from beta regression multilevel models.
(a) Main effect posterior densities for potential covariates of heavy-tailed dynamics. The response variable was Pr($\nu < 10$) (approximately the probability of heavy tails). The beta regression model was fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species. All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
(b) Taxonomic-order-level posterior densities of Pr($\nu < 10$) after accounting for time-series length. Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps). Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals). Short vertical line segments indicate median posterior estimates. Long dotted vertical line in panel b indicates the median expected Pr($\nu < 10$) from the prior distribution.
}
\label{fig:posteriors}
\end{center}
\end{figure}

>>>>

Black swans are statistically improbable events that nonetheless occur---often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two---a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$. We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}. For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for hierarchical intercepts at the taxonomic class, order, and species level (Supporting Information). We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}). However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example?)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans were usually downward events and were detected more frequently in longer time series and in populations with a smaller magnitude of process noise. Black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black-swan events were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics---particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors and were discarded (Supporting Information). A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. As an example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}. Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk---the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent). Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Fig.~\ref{fig:posteriors}b).
}
\label{fig:correlates}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{
Posterior probability distributions from beta regression multilevel models.
(a) Taxonomic-order-level posterior densities of Pr($\nu < 10$) (approximately the probability of heavy tails) after accounting for time-series length. Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps). Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals). Long dotted vertical line in panel a indicates the median expected Pr($\nu < 10$) from the prior distribution.
(b) Main effect posterior densities for potential covariates of Pr($\nu < 10$). The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species. All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
In both panels, short vertical line segments indicate median posterior estimates.
}
\label{fig:posteriors}
\end{center}
\end{figure}
